Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-16 ieee_paper.tex
Section 2 done.  Natasha's part.

@@

@@

In this paper, we propose yet a new idea, that is applicable to the
world of Finite-Difference Radial Basis functions (RBF-FD). A radial
basis functions refers to a set of shifted radially symmetric
functions $\phi_i(r) =\phi{||\xvec-\xvec_i||}$ centered at
$\xvec_i$\cite{}. Any function $f(\xvec)$ defined in a domain $\Omega$
can be expanded in this basis according to
\begin{equation}
f(\xvec) =  \sum_i w_i \phi_i(\xvec) \label{eq:rbf}
\end{equation}
% Wrong label eq:rbf in document 
Given function values at a set of discrete points $x_i$, one inverts
linear system (\ref{eq:rbf}) to obtain a set of weights $w_j$. Given a
set of $N$ RBF nodes, the matrix to invert is dense. Derivatives of
$f(\xvec)$ result from the differentiation of Eq. (\ref{eq:rbf}). For
very large problems, this is an computationally expensive since the
RBFs do not form an orthogonal basis, nor are they amenable to tensor
decomposition in the three coordinate directions. Rather, their
strength is the ability to randomly distribute nodes across complex
physical domains, and have an implementation that is independent of
dimensionality.

To alleviate the cost of a global approach, researchers \cite{} expand
$f(\xvec)$ in a local region of $x_i$. Thus,
$$
f(\xvec) =  \sum_{i=0}^{n_s} w_i \phi_i(\xvec) \label{eq:rbf}
$$
where one only considers a local set of $n_s$ neighbors of of point $x_j$. 
Finite-difference RBF expresses the derivative of $f(\xvec)$ at $x_j$ as a linear
combination of $f(\xvec)$ at the $n_s$ neighbors. Thus, 
$$
f = A w
$$ where $f$ is the vector of function values at the $n_s$ stencil
points (including the center point $x_j$. The matrix $A$ is a function
of $\phi_{ij}(\xvec) =\phi(||x_i-x_j||)$. This leads to the $n_s$
weight values $w = A^{-1} f$. The derivative of Eq. \ref{eq:rbf}
evaluated at $x_j$ gives
$$
   f'_j = \sum_{i=0}^{n_s} w_i \phi'_{ij}
$$
Substituting the vector of weights $w$ leads to 
$$
  f'_j = \sum_{i=0}^{n_s} \alpha_i f_i
$$
The derivative evaluation at all points of the domain $\Omega$, notwithstanding boundary conditions, takes the form
$$
  y = A x
$$ 
where $x$ is the souce vector of function values, and $y$ is the
vector of derivative values. The matrix $A$ is called a derivative
matrix. In the current litterature, the number of stencil points is a
fixed value, independent of the node under consideration\cite{}. The
computation of a single derivative has been reduced to a SpMV, where
each row has $n_s$ nonzeros. Because the sparsity of each row is
constant, ELLPACK~\cite{} becoes the most appropriate matrix
compression scheme\cite{}. In practice, $n_s=32$ in two-dimensional
flows and 64 or 100 for three-dimensional flows. These numbers are
similar to what is used in finite-element codes.

Many problems in fluid dynamics and in the geosciences require the solution to
transport equations of the form
$$
\pf{Q}{t} = f(Q,Q_x, Q_y, Q_z, \Laplacian{Q})
$$ 
where $Q$ is a vector of unknowns (3 components of velocity,
pressure, temperature). Thus, it is often necessary to compute the $x$
derivative of multiple functions, typically four for the Euler
equations, or five for the Navier-Stokes equations. Multiple
right-hand sides transform a SpMV into a SpMM (Sparse Matrix/dense
Matrix multiplication), which improves register utilization and
decreases cache misses by vectorizing over the multiple source
vectors. Further improvements are possible by recognizing that
different derivative matrices (x,y,z and Laplacian for example), have
the same sparsity distribution; only the weight change.  Thus,
alternative to computing a derivative of multiple functions, we can
calculate multiple derivatives of a single functions. The increased
memory bandwidth due to an increase in the number of derivaive
matrices is offset by better cache utilization, leading to an overall
benefit.

@@

\def\wide{2.5in}
\newfig{figures/rbf_stencils.pdf}{\wide}{RBFFD Stencils. Each node of the mesh is connected to $n_z-1$ stencil nodes in addition to itself. In the figure, node $A$ is connected to $B$, but $B$ is {\em not\/} connected to $A$. Thus adjacency graph of $A$ is not-symmetric. This must be corrected when reducing the bandwidth of $A$ with a Cuthill\-McKee algorithm, which assumes symmetry.}{fig:rbf_stencils}

@@

\begin{figure*}
\begin{verbatim}
const __m512i offsets = _mm512_set4_epi32(3,2,1,0);
const __m512i four = _mm512_set4_epi32(4,4,4,4); 
int int_mask = 0x1111;
__mmask16 mask = _mm512_int2mask(int_mask);

@@

    //three more times with BBBB, CCCC, and DDDD
  }
  _mm512_storenrngo_ps(dom.result_vt+nb_mat*nb_vec*r, accu);  
} 
\end{verbatim}
\caption{mat mul code.}
\label{code:mat_mul}
\end{figure*}

@@

  \subfigure[Performance of $y=Ax$ on the MIC. Squares: base 1/1 case,
    solid circles: 4/4 case implemented in C++, solid triangles: 4/4
    case implemented with MIC vector instructions. $-O3$ compilation
    options. Each group consists of four cases: grids of $64^3$ and
    $96^3$ with and without Reverse Cuthill
    McKee.\todo{I don't think this is supercompact but RBF 3D.}]{\includegraphics[width=.49\linewidth]{figures/mic_performance_nb_threads.pdf}\label{}}
%
  \subfigure[performance on the host on a 64x64x64 matrix. Manual
    vectoization brings significant improvement there as
    well.]{\includegraphics[width=.49\linewidth]{figures/64x64x64.pdf}\label{}}

  \caption{Vectorization matters in practice. Differences in BW
    translate in different performance.\todo{this two plots should ``correlate'' }}
\end{figure*}

@@

\section*{Acknowledgment}
The first and last authors acknowledge NSF funding under NSF grant DMS-\#0934331 (FSU).

>>>>

\usepackage{listings}

@@

\usepackage{amsmath}
\usepackage{amssymb}

@@

In this paper, we propose yet a new idea that is applicable to radial basis functions (RBFs). Their strength is the ability to randomly distribute points across complex physical domains, and have an implementation that is independent of dimensionality. RBFs approximate a function $f(\xvec)\subset \mathbb{R}^d$ sampled at a set of $N$ distinct point locations, $x_j$, by linearly combining translates of a single radially symmetric function $\phi(r)$, where  $r = \|\xvec-\xvec_{j}\|$ denotes the Euclidean distance (e.g., in 2-D $\sqrt{(x-x_j)^2+(y-y_j)^2}$) between where the function is evaluated $\xvec$ and where the RBF is centered $\xvec_{j}$. That is, the interpolant is $s(\xvec) =  \sum_{j=1}^{N} w_j \phi_i(\|\xvec-\xvec_{j}\|)$. The weights $w_j$ are obtained by inverting the system
\begin{equation}
\parray{lccr}{
\phi_{11} & \phi_{12} & \cdots & \phi_{1N} \\
\vdots & \ddots & \vdots & \vdots \\
\phi_{n1} & \phi_{n2} & \cdots & \phi_{NN} 
}
\parray{c}{ w_{1} \\ \vdots \\ w_{N} }
=
\parray{c}{ f(\xvec_1) \\ \vdots\\ f(\xvec_N) }. 
\label{eq:rbf}
\end{equation}
where $\phi_{ij} = \phi(\|\xvec_i-\xvec_j\|)$. 
%\end{equation}
%\begin{equation}
%\begin{bmatrix}
%\phi(\|\xvec_{1}-\xvec_{1}\|) & \phi(\|\xvec_{1}-\xvec_{2}\|) & \cdots & \phi(\|\xvec_{1}-\xvec_{N}\|) \\
%\vdots & \ddots & \vdots  \\
%\phi (\|\xvec_{n}-\xvec_{1}\|) & \phi(\|\xvec_{n}-\xvec_{2}\|) & \cdots & \phi(\|\xvec_{N}-\xvec_{N}\|)
%\end{bmatrix}

The RBF differentiation matrix, $D_N$, is derived by applying the desired analytic derivative operator $L$ to the RBF interpolant $s(\xvec)$ above and evaluating it at the point locations. For very large problems, this is a computationally expensive since the matrix in (\ref{eq:rbf}) is full and inversion requires O$(N^3)$ operations. To alleviate the cost of this global approach (i.e. using every node in the domain to calculate the derivative at a given node $\xvec_i$), RBF-generated finite differences (RBF-FD) have been derived \cite{TAI1,TAI2,SDY02,WrFo06,FoL11,FLBWSC12}. RBF-FD use only a local set of the $n_z-1$ nearest neighbors to the point $\xvec_i$ to approximate the derivative. In other words, $Lf(\xvec_i)=\sum_{j=1}^{n_z}a_jf(\xvec_j)$. The differentiation weights, $a_j$, are calculated by enforcing that this linear combination should be exact for RBFs, $\{\phi(\|\xvec-\xvec_{j}\|)\}_{j=1}^n_z$, centered at each of the node locations $\{\xvec_j\}_{j=1}^{n_z}$ (classical finite differences (FD) would enforce that it be exact for polynomials instead). Similar to FD, as the stencil size $n_z$ increases so does the order of the method.

For a total of $N$ points, there will be $N$ linear systems to solve, each of size $n_z \times n_z$. Each linear solve produces a row of the RBF-FD differentiation matrix $D_{n_z}$, resulting in a $N \ times N$ matrix with $n_zN$ nonzero entries. To evaluate the derivative at all points in the domain, takes the form
$$
  \mathbf{g} = D_{n_z} \mathbf{f}
$$
where $f$ is the source vector of function values and $g$ is the resulting vector of derivative values. The computation of a single derivative has been reduced to a SpMV, where each row has $n_z$ nonzeros. Since the sparsity of each row is constant, ELLPACK~\cite{} becomes the most appropriate matrix compression scheme\cite{}. In practice, $n_z=32$ in two-dimensional flows and 64 or 100 for three-dimensional flows. These numbers are similar to what is used in finite-element codes.

Many problems in fluid dynamics and in the geosciences require the solution to transport equations of the form
$$
\pf{Q}{t} = f(Q,Q_x, Q_y, Q_z, \Laplacian{Q})
$$
where $Q$ is a vector of unknowns (3 components of velocity,
pressure, temperature). For example, in solving a system of equations, it is often necessary to compute the $x$
derivative of multiple functions, typically four for the Euler
equations, or five for the Navier-Stokes equations. Multiple
right-hand sides transform a SpMV into a SpMM (Sparse Matrix/dense
Matrix multiplication), which improves register utilization and
decreases cache misses by vectorizing over the multiple source
vectors. Further improvements are possible by recognizing that
different derivative matrices (x,y,z and Laplacian for example), have
the same sparsity distribution; only the weight change.  Thus,
alternative to computing a derivative of multiple functions, we can
calculate multiple derivatives of a single functions. The increased
memory bandwidth due to an increase in the number of derivative
matrices is offset by better cache utilization, leading to an overall
benefit.

% GORDON I WOULD ELIMINATE BELOW AND END HERE.

@@

\def\wide{2.5in}
\newfig{figures/rbf_stencils.pdf}{\wide}{RBFFD Stencils. Each node of the stencil is connected to $n_z-1$ stencil nodes in addition to itself. In the figure, node $A$ is connected to $B$, but $B$ is {\em not\/} connected to $A$. Thus adjacency graph of $A$ is not-symmetric. This must be corrected when reducing the bandwidth of $A$ with a Cuthill\-McKee algorithm, which assumes symmetry.}{fig:rbf_stencils}

@@

\begin{figure}[h]
\begin{center}
\lstset{language=c++,basicstyle=\scriptsize,numbers=left,numbersep=5pt}
\begin{lstlisting}{float=h, Caption}
const __m512i offsets = _mm512_set4_epi32(3,2,1,0);
const __m512i four = _mm512_set4_epi32(4,4,4,4); 
int int_mask = 0x1111;
__mmask16 mask = _mm512_int2mask(int_mask);

@@

    //three more times with BBBB, CCCC, and DDDD
  }
  _mm512_storenrngo_ps(dom.result_vt+nb_mat*nb_vec*r, accu);  
\end{lstlisting}
\end{center}
\end{figure}

%\begin{figure*}
%\begin{verbatim}
%const __m512i offsets = _mm512_set4_epi32(3,2,1,0);
%const __m512i four = _mm512_set4_epi32(4,4,4,4); 
%int int_mask = 0x1111;
%__mmask16 mask = _mm512_int2mask(int_mask);
%
%#pragma omp parallel for
%for (int r=0; r < nb_rows; r++) {
%  __m512 accu = _mm512_setzero_ps();
%
%  for (int n=0; n < nz; n+=4) {
%    float* a = &dom.col_id_t[0]+n+nz*r;
%    __m512i vect_i = _mm512_setzero_ps();
%    vect_i = _mm512_mask_loadunpacklo_ps(vect_i, mask, a);
%    vect_i = _mm512_mask_loadunpackhi_ps(vect_i, mask, a);
%    vect_i = _mm512_swizzle_ps(vect_i, _MM_SWIZ_REG_AAAA);
%    vect_i = _mm512_fmadd_epi32(vect_i, four, offsets);
%
%    all_vect = _mm512_i32gather_ps(vect_i, dom.vec_vt, scale);
%
%    mat_ent = _mm512_load_ps(dom.data_t + nb_mat*(n + r*nz));
%
%    //perform first tensor multiplication
%    correct_vect = permute(all_vect, _MM_PERM_AAAA);
%    correct_mat = _mm512_swizzle_ps(mat_ent, _MM_SWIZ_REG_AAAA);
%    accu = _mm512_fmadd_ps(correct_vect, correct_mat, accu);
%
%    //three more times with BBBB, CCCC, and DDDD
%  }
%  _mm512_storenrngo_ps(dom.result_vt+nb_mat*nb_vec*r, accu);  
%} 
%\end{verbatim}
%\caption{mat mul code.}
%\label{code:mat_mul}
%\end{figure*}

@@

  \subfigure[Performance of $y=Ax$ on the MIC. Squares: base 1/1 case,
    solid circles: 4/4 case implemented in C++, solid triangles: 4/4
    case implemented with MIC vector instructions. $-O3$ compilation
    options. Each symbol has two cases: with (red) and without blue) 
    with and without Reverse Cuthill-McKee. The grid is $96^3$.}
    {\includegraphics[width=.49\linewidth]{figures/mic_performance_nb_threads.pdf}\label{}}
%
  \subfigure[performance on the host on a sparse matrix with $64^3$ rows. Manual
    vectoization brings significant improvement there as
    well.]{\includegraphics[width=.49\linewidth]{figures/64x64x64.pdf}\label{}}
  \caption{Vectorization matters in practice. Differences in BW
    translate in different performance.\todo{this two plots should ``correlate'' }}
\end{figure*}

@@

\section*{Acknowledgment}
The first and last authors acknowledge NSF funding under NSF grant DMS-\#0934331 (FSU). 
NCAR is sponsored by the National Science Foundation. N. Flyer      
acknowledges support of NSF grant DMS-0934317.
